Race Science in Mainstream Psychology, 1960-2010

### 2024-05-03 paper/paper.qmd
prep to merge back into master

---
title: "Race Science in Mainstream Psychology, 1960-2010"
# authors:
#   - name: "Daniel J. Hicks"
#     email: "dhicks4@ucmerced.edu"
#     affiliations: 
#       - id: ucm
#         name: "University of California, Merced"
#         city: "Merced"
#         country: "CA, USA"
#     corresponding: true
#   - name: "Emilio J. C. Lobato"
#     affiliations:
#       - ref: ucm
abstract: |
  | While anthropologists and human geneticists have long rejected the biological race concept, both social movements and prominent researchers have purported to justify racial inequality and colonialism by appealing to the epistemic authority of science.
  In this study, bibliometric and text mining methods were used to trace the appearance of race science discourse in both mainstream behavioral science journals and the parascholarly journal *Mankind Quarterly* from 1960 to 2010. 
  Our analysis found two distinguishable lines of race science discourse.  One is strongly associated with *Mankind Quarterly*, while the other — emphasizing race and intelligence — is strongly associated with mainstream journals, especially *Intelligence* and *Personality and Individual Differences*.  
  Psychology, and especially intelligence research, has played a significant role in mainstreaming race science.

@@

bibliography: [references.bib, refs-manual.bib]
cite-method: "citeproc"
csl: "nature-no-superscript.csl"

format: 
  elsevier-pdf:
    journal: 
      name: "journal"
    keep_tex: true
    latex_engine: "xelatex"
    number-sections: false
    link-citations: true
    mathspec: true

@@

@@

@@

![Journals publishing 6 or more Pioneer-funded researchers, Web of Science author search results.](img/wos_results.png){#fig-wos width="4.76in" height="2.6in"}

@@

![Silge plots [@SilgeTopicModeling2017] and smoothed time series for three focal topics.  Top row: top 15 terms by $\beta$ (term-topic distribution) for each topic.  Bottom row: count of articles associated with each topic, by journal and year.  Article counts use a threshold approach, with $\gamma$ (topic-document distribution) greater than 0.5.  Thin lines give annual values, thick lines give 5-year running averages.  Because *Behavior Genetics* and *Behavioral and Brain Sciences* are not prominent in any panel, neither is given a direct label.](img/focal_topics.png){#fig-focal width="4.76in" height="3.57in"}

@@

![Corpus assembly.](img/dataset.png){#fig-corpus width="120%"}

@@

![Count of documents in the corpus, by journal and year.](img/02_count.png){#fig-counts width="4.76in" height="3.6in"}

@@

@@

{{< include ../scripts/11-proof-of-concept.qmd >}}

@@

>>>>

---
title: "Race Science in Mainstream Psychology, 1960-2010"
authors:
  - name: "Daniel J. Hicks"
    email: "dhicks4@ucmerced.edu"
    affiliations: 
      - id: a
        name: "University of California, Merced"
        city: "Merced"
        country: "CA, USA"
    corresponding: true
  - name: "Emilio J. C. Lobato"
    affiliations:
      - ref: a

abstract: |
  | While anthropologists and human geneticists have long rejected the biological race concept, both social movements and prominent researchers have purported to justify racial inequality and colonialism by appealing to the epistemic authority of science.
  In this study, bibliometric and text mining methods were used to trace the appearance of race science discourse in both mainstream behavioral science journals and the parascholarly journal *Mankind Quarterly* from 1960 to 2010. 
  Our analysis found two distinguishable lines of race science discourse.  One is strongly associated with *Mankind Quarterly*, while the other — emphasizing race and intelligence — is strongly associated with mainstream journals, especially *Intelligence* and *Personality and Individual Differences*.  
  Psychology, and especially intelligence research, has played a significant role in mainstreaming race science.

@@

bibliography: [references.bib, refs-manual.bib]
# cite-method: "citeproc"
# csl: "nature-no-superscript.csl"

format: 
  pnas-pdf:
    keep_tex: true
    latex_engine: "xelatex"
    link-citations: true
    mathspec: true
    classoption:
      - twocolumn
      - twoside
    fig.env: "figure*"

@@

<!-- 
PNAS guidelines
Research Report/Direct Submission: The preferred length of these articles is 6 pages, but PNAS allows articles up to a maximum of 12 pages. A standard 6-page article is approximately 4,000 words, 50 references, and 4 medium-size graphical elements (i.e., figures and tables).

- a manuscript file (in any format) including the following:
    title page (title, author list, classification, keywords)
    abstract
    significance statement
    main text
    references
    figures or tables with appropriate legends (may be uploaded separately)
    SI files (may be uploaded separately)
- contact and competing interest information for all authors
- data sharing plans (for all data, documentation, and code used in analysis)
- funding information and whether an open access license has been selected
- list of appropriate Editorial Board members, NAS members, and qualified reviewers (minimum of three each) who are experts in the paper's scientific area. (A brief justification for suggested reviewers is welcome, particularly for interdisciplinary papers)

- revised submission will require track changes version

figure sizes: 
Small: approximately 9 cm x 6 cm
Medium: approximately 11 cm x 11 cm
Large: approximately 18 cm x 22 cm

 -->

*[TODO: 
- NHB reviews todo
- sort git tree
- reorganize into PNAS ms + SI
- check cite formatting
- copy updated plots? 
- abstract formatting
- figure sizes
]*

@@

\onecolumn

@@

\twocolumn

![Journals publishing 6 or more Pioneer-funded researchers, Web of Science author search results.](img/wos_results.png){#fig-wos width="4.76in" height="2.6in" fig-env="figure*"}

@@

![Silge plots [@SilgeTopicModeling2017] and smoothed time series for three focal topics.  Top row: top 15 terms by $\beta$ (term-topic distribution) for each topic.  Bottom row: count of articles associated with each topic, by journal and year.  Article counts use a threshold approach, with $\gamma$ (topic-document distribution) greater than 0.5.  Thin lines give annual values, thick lines give 5-year running averages.  Because *Behavior Genetics* and *Behavioral and Brain Sciences* are not prominent in any panel, neither is given a direct label.](img/focal_topics.png){#fig-focal width="4.76in" height="3.57in" fig-env="figure*"}

@@

![Corpus assembly.](img/dataset.png){#fig-corpus width="120%" fig-env="figure*"}

@@

![Count of documents in the corpus, by journal and year.](img/02_count.png){#fig-counts width="11cm" height="11cm" fig-env="figure*"}

\onecolumn

@@

\twocolumn

@@

<!-- {{< include ../scripts/11-proof-of-concept.qmd >}} -->

@@

### STM robustness check

As a robustness check, we compared the `tmfast` results with a topic model fit using `STM`, a widely used package for topic modeling in R *[cite]*.  Using the medium vocabulary, $k=40$, and a relaxed convergence threshold (`1e-4` compared to the default `1e-5`), fitting a single `STM` model took approximately 20 minutes.  This was approximately two orders of magnitude slower than `tmfast`.  And, unlike `tmfast`, `STM` can only fit a single value of $k$ at a time.  We therefore only fit this one `STM` model.

*[xref]* shows four topics extracted from the `STM` model, automatically identified by the use of race keywords, intelligence keywords, or both, among the top 15 terms.  As in the `tmfast` analysis, the "race only" topic is almost exclusive to MQ; the "race and intelligence" topic emerges in mainstream journals before appearing in both the mainstream and MQ; and there are distinctive "intelligence only" topics.  Compare *[xref]*.  One notable difference between these results is the role of PR in publishing race and intelligence.

![](img/12-stm.png){fig-env="figure*"}

### *[author analysis]*

### 2024-05-03 paper/test.qmd
prep to merge back into master

## Simulation parameters
k = 20            ## total number of topics
N = 100           ## papers per journal
vocab = 1000      ## vocabulary size
paper_size = 500  ## length of each paper
```

We construct a simulated corpus as follows.  We work with five journals, labelled A through E.  Journal A primarily publishes the marginalized topic (correspond to race science).  Ten journals, C-L, are mainstream journals, and publish an even mix from `r k-1` other topics, but avoid the marginalized topic (weighted 0 in the Dirichlet parameterization below).  These journals represent the context or comparison group for identifying the race science topic; their exact number is not important here.

Journal B is the venue by which the marginalized topic is "mainstreamed."  This journal is represented in a time-indexed way, $B_t$, with $t$ ranging from 0 to 5.  For simplicity, in the text each time-slice of B is treated as a distinct journal, giving us 17 journals total (A; the ten C-L; plus 6 time-slices of B).

This gives us three sets of journals: A, B (0-5), and C-L.  In the LDA model, these journal sets correspond to $\mathbf{\alpha}$, the parameter for the "prior" Dirichlet distribution $\Theta(\mathbf{\alpha})$. The `tmfast` package provides a helper function `peak_alpha()` for constructing $\mathbf{\alpha}$.

```{r}
## Construct journal theta ----
## A: marginal topic journal
A_alpha = peak_alpha(k, 1, peak = .9)
A_alpha
## B: mainstream journal that picks up marginal topic over time
B_alpha = map(0:5, ~peak_alpha(k, 1, peak = 0.05*.x)) |> 
    set_names(glue('B{0:5}'))
## C-E: mainstream journals
C_alpha = rep.int(list(peak_alpha(k, 1, peak = 0)), 10) |> 
    set_names(LETTERS[3:12])
C_alpha$C

alphas = c(list(A = A_alpha), 
           B_alpha, 
           C_alpha)

## 17 journals/journals-at-time-t
length(alphas)
```

Each paper has a simplex $\theta \sim \Theta(\mathbf{\alpha})$, representing a distribution over topics. We draw `r N` papers from each journal.

```{r}
theta = map(alphas, ~ rdirichlet(N, .x)) |> 
    imap(~ magrittr::set_rownames(.x, glue('{.y}-{str_pad(1:N, 
                                           2, 
                                           pad = "0")}'))) |> 
    reduce(rbind)

head(theta)

theta_df = theta |> 
    as_tibble(rownames = 'doc', 
              .name_repair = tmfast:::make_colnames) |> 
    pivot_longer(cols = -doc, 
                 names_to = 'topic', 
                 values_to = 'theta')

theta_df
```

We next define $\phi$, the distribution over words for each topic.  We use a vocabulary of `r vocab` words, and draw each topic's $\phi$ as a simplex from a symmetric Dirichlet distribution with $\alpha = 0.1$. Note that words are simply natural numbers.

```{r}
## Word distributions ----
## 20 topics, 1000 terms, 
## uniform Dirichlet distribution w/ param 0.1
phi = rdirichlet(k, 0.1, k = vocab)
str(phi)
```

For simplicity, we use a uniform length of `r paper_size` words for each paper. `draw_corpus` assumes that this is a vector with one element for each document.

```{r}
## Document length ----
## 500 words each
doc_len = rep.int(paper_size, N * length(alphas))
```

With these elements constructed, `draw_corpus()` generates a synthetic document-term matrix. Because this process is relatively slow (approximately 75 seconds on the laptop used to prepare this simulation), we use `memoise` to create a cache. As a pre-processing step for the topic model, we apply the `log(x+1)` transform.

```{r}
## Draw corpus ----
draw_corpus_cached = memoise(draw_corpus, 
                              cache = cachem::cache_disk('draw_corpus_cache'))
set.seed(2024-03-25)  ## NB draw_corpus() is non-deterministic! 
tic()
corpus = draw_corpus_cached(doc_len, theta, phi) |> 
    mutate(doc = rownames(theta)[doc], 
           n = log1p(n))
toc()
corpus
```

We can now fit the topic model.

```{r}
## Fit topic model ----
fitted = tmfast(corpus, c(10, 15, 20, 25, 30))
```

The tile plot indicates that (1) documents in journal A generally have a high probability for topic 1; journals C-E generally have a low probability; and journal B gradually has more articles in topic 1.

```{r}
## Topics by journal ----
add_journal = function(x) {
    mutate(x, journal = str_extract(document, '^[A-Z]*[0-9]?'))
}

tidy(fitted, k = 20, matrix = 'gamma') |> 
    add_journal() |> 
    ggplot(aes(document, topic, fill = gamma)) +
    geom_tile() +
    facet_wrap(vars(journal), scales = 'free', 
               ncol = 7) +
    scale_x_discrete(guide = 'none')
```

Focusing on topic 1, we first calculate the average probability across each journal.  A has the highest value; C-E are consistently low; and B shows an increase over time.

```{r}
## Topic model clearly detects the focal topic in A, increase in B over time, minimal presence in other journals
tidy(fitted, k = 20, matrix = 'gamma') |> 
    filter(topic == 'V01') |> 
    add_journal() |> 
    ggplot(aes(journal, gamma)) +
    ggforce::geom_sina() +
    stat_summary(color = 'red')
```

While the journals are largely distinguishable on the other topics, except for the unusually low value for A. 
```{r}
## With no pattern for the other topics
tidy(fitted, k = 20, matrix = 'gamma') |> 
    filter(topic == 'V02') |> 
    add_journal() |> 
    ggplot(aes(journal, gamma)) +
    ggforce::geom_sina() +
    stat_summary(color = 'red')
```

Next we calculate the rate of documents with probability greater than 0.1 in topic 1. As expected, this is almost all papers in journal A, no papers in the other mainstream journals, and in B the rate increases over time.

```{r}
tidy(fitted, k = 20, matrix = 'gamma') |> 
    filter(topic == 'V01', 
           gamma > .1) |> 
    add_journal() |> 
    count(journal) |> 
    mutate(rate = n / N)
```

We therefore conclude that this topic model method can detect the "mainstreaming" from a marginal topic over time.

As a limitation, we note that, for $\theta > 0.05$ or so, the estimated topic probabilities are below the true values. 
```{r}
tidy(fitted, k = 20, matrix = 'gamma') |> 
    filter(topic == 'V01') |> 
    add_journal() |> 
    inner_join(theta_df, by = c('document' = 'doc', 'topic')) |> 
    ggplot(aes(theta, gamma, color = journal)) +
    geom_point() +
    stat_function(fun = identity, 
                  inherit.aes = FALSE, color = 'black') +
    scale_color_viridis_d()
```

*[tmfast paper]* notes this problem, and implements a "renormalization" option when extracting the probability distributions from the fitted model.  This "renormalization" increases the "peakedness" of the distribution (lower entropy), adjusting the highest probabilities up and making the small probabilities several orders of magnitude smaller.  Renormalization would just amplify the differences found here, and so is not implemented for this proof-of-concept.  Renormalization is used in the primary, empirical analysis.

>>>>

## Simulation parameters
k = 20            ## total number of topics
N = 100           ## papers per journal
vocab = 1000      ## vocabulary size
paper_size = 500  ## length of each paper

paper_size
```
